---
title: 'Detecting Frequent Words in UK Manifestos'
output: distill::distill_article
---

By: Akmal G on March 26th, 2026

In this project, I looked at the use of diction and the frequency of certain words, in order to deduce speech patterns of UK politicians, using evidence from a corpus of manifestos. 

I first load the following packages; quanteda, readtext, ggplot2, ggthemes and tidyverse.

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}
library(quanteda)
library(readtext)
library(ggplot2)
library(ggthemes)
library(tidyverse)
```

The argument docvarsfrom= allows us to use the file names as document ids; '> corpusSource object containing 69 texts of manifestos and 1 docvar'. We then need to turn the text files into a 'corpus', just to make sure the docnames are as expected (is being corrected in rebuild). Using ntoken and ntype, we can give names to the documents in question.

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}
manifestos <- readtext("UK Manifestos/*.txt", docvarsfrom = c("filenames"))
manifestos #to check
manifestos_corpus <- corpus(manifestos, docnames = manifestos$doc_id)

docnames(manifestos_corpus)
ntk <- ntoken(manifestos_corpus) # shows the tokens per document
ntps <- ntype(manifestos_corpus) # shows the types per document

# There are 562339 tokens per document and 121676 types per document.
sum(ntk)
sum(ntps)

```

There is a difference  between types and tokens.To summarize the corpus, for example, we may want to consider whether UK party manifestos getting longer or shorter over time. We can look at the number of sentences for each manifesto. I grabbed the dates of the manifestos, which could be done with docvars()<, but could also be done with regex.

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}
num_sentences <- nsentence(manifestos_corpus)
num_sentences
sum(num_sentences)

nms <- names (manifestos_corpus) # document names
head(nms)

# The following code extracts the digits (in form of dates) from the document names
Dates <- as.numeric(gsub('[^[:digit:]]','', nms))
head(Dates)

```

We notice from sum(num_sentences) that there 24524 words to look through.Then we plot them either in base scatterplot as well as a nicer ggplot.

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}
plot(Dates, num_sentences, pch=16, ylab = "Number of Sentences", xlab = "Year")

info_summary <- data.frame(Dates = Dates, Sentences = num_sentences)
head(info_summary)
info_summary[69,]
ggplot(data=info_summary, aes(x=Dates, y=Sentences, group=1)) +  geom_point() +    
  theme_hc()
```

## Looking inside the Documents 

To see this, lets grab the Labor 1983 manifesto alone, then we transform them into tokens. Simply using the length function counts the number of tokens to around 25000; using the unique funtion counts the number of unique tokens to around 3600

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}
Lab_1983 <- corpus_subset(manifestos_corpus, docvar1=="Lab1983")
summary(Lab_1983)

tokens_all <- tokens(Lab_1983) #this defaults to the /word/ level
head(tokens_all)

length(tokens_all[[1]]) #~25k
length(unique(tokens_all[[1]])) #~3.6k
```
For grater clarity, we could remove punctuation and some numbers. 
Note that your research question should drive this choice and length of your documents might play a role. 

More Information: [DennySpirling](https://www.cambridge.org/core/journals/political-analysis/article/abs/text-preprocessing-for-unsupervised-learning-why-it-matters-when-it-misleads-and-what-to-do-about-it/AA7D4DE0AA6AB208502515AE3EC6989E)

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}
# Check and remove punctuation tokens; then place it in an object.
tokens_nopunc <- tokens(Lab_1983, remove_punct=T)
length(tokens_nopunc[[1]]) #~22.5k

# Check and remove punctuation tokens as well as numbers.
tokens_nopunc_nonum <- tokens(Lab_1983, remove_punct=T, remove_number=T)
length(tokens_nopunc_nonum[[1]]) #~22.4k
```

It would help us to separated bi-grams (only) from a set of uni-grams and bi-grams. 
While tokens_ngrams() generates n-grams or skip-grams in all possible combinations of tokens, tokens_compound() generates n-grams more selectively. Bi-grams can be negated using phrase() and a wild card (*).

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}
toks_ngram <- tokens_ngrams(tokens_nopunc_nonum, n = 2)
head(toks_ngram[[1]], 20)
# Reveal the top part of the list of bi-grams.

toks_neg_bigram <- tokens_compound(tokens_nopunc_nonum, pattern = phrase("not *"))
head(toks_neg_bigram[[1]], 20)
# Reveal the top part of the list of uni-grams - notice the lack of bi-grams.

toks_neg_bigram_select <- tokens_select(toks_neg_bigram, pattern = phrase("not_*"))
head(toks_neg_bigram_select[[1]], 20)
# Reveal the top part of the list of bi-grams with 'not' in front.

```

The material in Quanteda recommends using the *tokens_compound* command to get top tokens rather than the *tokens_ngrams*. 

## Stemming 

Using the Porter Stemmer, I can see what package this is coming from an what stemmed it is using by doing ?tokens_wordstem. Stemming removes all words of the similar family, in this case, the word 'family'.

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}
stemmed <- tokens_wordstem(tokens_all) #can inspect directly, if desired.
# to see what's happened in practice, let's compare 'family' and 'families'.
# So, compare
tokens_all[[1]][grep("fam",tokens_all[[1]])]
# to
stemmed[[1]][grep("fam",stemmed[[1]])]
```

## Stopwords

I would also like to remove the all stopwords, in English.  

```{r, message = FALSE, error = FALSE, warning = FALSE}
stopwords("english")
```

## Vector Space Representation 

```{r, results = 'hide', echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}
Lab_1983 <- corpus_subset(manifestos_corpus, docvar1=="Lab1983")
Lab1983_vector <- tokens(Lab_1983)
# Categorize any words into any available tokens.
Lab1983_vector <- tokens_wordstem(Lab1983_vector)
# Categorize tokens into word steams
Lab1983_vector <- dfm(Lab1983_vector, tolower = TRUE)
# Transform data into readable matrix
head(Lab1983_vector)
```

With the following objects, I can make a dfm of the whole Manifestos collection, and give weight to the most common words in the corpus.

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}

# Create object with all manifestos for past UK elections.
manifestos <- readtext("UK Manifestos/*.txt", docvarsfrom = c("filenames"))
manifestos <- corpus(manifestos)

# Remove all punctuation and English stopwords from each speech.
dfmat_man <- tokens(manifestos, remove_punct = TRUE) |>
  tokens_remove(stopwords("en")) |>
  dfm()
dfmat_man

# Create an alternate weighted set using the frequency-inverse document frequency (tf-idf).
dfmat_man_weighted <- dfm_tfidf(dfmat_man)

# Compare and contrast dfmat_man and dfmat_man_weighted
topfeatures(dfmat_man, 30)
topfeatures(dfmat_man_weighted, 30)

```

## Visual Representation of Word Frequency

We can visualize this with ggplot's bar graphs. It better shows the frequency of words in dfmat_man. We mutate the data for dfmat_man, then we visualize it. The object dfmat_man, on one hand, provides a sum of the frequency of certain keywords. This includes 'government', 'new', 'people', 'must', 'can' and 'local', which may initially indicate the frequency of how candidates tend to provide election promises at their incumbencies. The function dfmat_man_weighted, on the other hand, gives these words  weight in comparison to their frequency among the whole group; judging on  rare or frequent they may appear.

For dfmat_man;

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}
top_raw <- topfeatures(dfmat_man, 20) %>% 
  as.data.frame() %>% 
  rownames_to_column("word") %>% 
  rename(count = ".")

ggplot(top_raw, aes(x = reorder(word, count), y = count)) +
  geom_col(fill = "salmon") +
  coord_flip() + # Makes it a horizontal bar chart so words are readable
  theme_minimal() +
  labs(title = "Top 20 Most Frequent Words in UK Manifestors", x = NULL, y = "Raw Count")
```
For dfmat_man_weighted;

```{r, echo=TRUE, message = FALSE, error = FALSE, warning = FALSE}

top_weighted <- topfeatures(dfmat_man_weighted, 20) %>% 
  as.data.frame() %>% 
  rownames_to_column("word") %>% 
  rename(tfidf = ".")

ggplot(top_weighted, aes(x = reorder(word, tfidf), y = tfidf)) +
  geom_col(fill = "palevioletred") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 20 Most Unique Words, Mutated with TF-IDF", x = NULL, y = "TF-IDF Weight")

```


