---
title: "Societal Determinism: How do Mothers have an Influence on a Child's IQ?"
output: distill::distill_article
---

By: Akmal G on November 4th, 2025

For one of my first works while learning from the University of Essex, I created a visualization that shows how a mother's education, IQ and work status combined to influence children' test scores. The visualization revealed a number of patterns of advantages and disadvantages across different family backgrounds.

The kidiq data stems from the National Longitudinal Survey of Youth and is used in Gelman and Hill (2007). The data set contains 434 observations measured on five variables, and is obtained from https: //github.com/jknowles/BDAexampleR.data

I used the packages tidyverse, stringr, readr, ggplot2 and curl.

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(readr)
library(ggplot2)
library(curl)
```

```{r setup but tidyverse, include=FALSE}
library(tidyverse)
```

My approach was to create categories that combine mother's education and work status. Then, I processed child outcomes vary across these combined categories. 

```{r, create categories}
# Download and mutate data
kidiq.raw <- read_csv("kidiq-2.csv", show_col_types = FALSE)
kidiq <- kidiq.raw %>%
mutate(
  
# Create work status labels
work_status = case_when(
mom_work == 1 ~ "Not working",
mom_work == 2 ~ "Part-time",
mom_work == 3 ~ "Full-time",
mom_work == 4 ~ "Full-time+"
),
# Create highschool diploma success groups
highschool_diploma = case_when(
mom_hs == 0 ~ "High School Diploma",
mom_hs == 1 ~ "No High School Diploma",
),
# Create combined advantage categories
family_advantage = case_when(
mom_hs == 0 & mom_work <= 2 ~ "Low",
mom_hs == 0 & mom_work > 2 ~ "Mixed",
mom_hs == 1 & mom_work <= 2 ~ "Mixed",
mom_hs == 1 & mom_work > 2 ~ "High",
),
# Create age groups
age_group = case_when(
mom_age >= 17 & mom_age <= 20 ~ "Ages 17-20",
mom_age >= 21 & mom_age <= 23 ~ "Ages 21-23",
mom_age >= 24 & mom_age <= 26 ~ "Ages 24-26",
mom_age >= 27 & mom_age <= 29 ~ "Ages 27-29"
),
# Create IQ terciles for grouping
iq_group = case_when(
mom_iq <= quantile(mom_iq, 0.33, na.rm = TRUE) ~ "Lower IQ",
mom_iq <= quantile(mom_iq, 0.67, na.rm = TRUE) ~ "Middle IQ",
TRUE ~ "Higher IQ"
)
)

kidiq$age_group <- factor(kidiq$age_group, labels = c(
"Ages 17-20","Ages 21-23", "Ages 24-26","Ages 27-29"))
kidiq$iq_group <- factor(kidiq$iq_group, labels = c(
"Lower IQ", "Middle IQ", "Higher IQ"))
kidiq$work_status <- factor(kidiq$work_status, labels = c(
"Not working", "Part-time", "Full-time","Full-time+"))
kidiq$family_advantage <- factor(kidiq$family_advantage, labels = c(
"Low", "Mixed", "High"))
kidiq <- as.data.frame(kidiq)
```

To indicate different effects on IQ, I added additional dimensions using different colors, size, and faceting to differentiate. In the end, I was able to tell a story about social mobility and opportunity through a readable violin plot.

```{r, create violin}
k1 <- ggplot(kidiq, aes(x = family_advantage, y = kid_score, color = age_group)) +
geom_violin() +
facet_wrap(~age_group) +
scale_color_brewer(palette = "RdPu") +
labs(
title = "Family Ties? Does a mother’s educational and employment
background affect their child's performance",
subtitle = "Correlation between family advantage with child's",
x = "Mother's Household Advantage",
y = "Child's Test Score",
caption = "Data: Test scores and mother's characteristics") +
theme_minimal()
print(k1)
```

The correlation between a mother's educational and employment background, collectively known as Family Advantage, and the academic performance of their children is shown across several age groups. Most data seems to be sporadic,with notable exceptions being unemployed and full-time working mothers in the  age range 24-26, and unemployed mothers in their late teens, having children who reached the mode performance scores.


